Figures to insert in text

### 2018-03-16 Chapters/Chapter02.tex
SDM more grammar and typo fixes

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. This model may be applied in all sort of applications because of its incredible ability to closely reflect the human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where one's fear comes from. One just feels it. We may interpret this situation as clues of the present --- a dark alley; a giant metropolitan area; people going on about their lives mostly indifferent from each other; constrained routes ahead and behind you; a general lack of activity; that very feeling that something isn't right... without knowing precisely what isn't right, or even what `right' means... etc. --- recalling past experiences from memory and thus generating the feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, one's brain involuntarily makes an analogy and recognizes the possibility of danger. This flexibility into mapping one situation in another is an important human feature which is hard to replicate in computers.

@@

Our motivation was our own effort to run our models. As there is no reference implementation, we had to implement our own and run several simulations to ensure that our implementation was correct and bug-free. Thus, we had to deviate from our main goal --- which was to test our hypothesis and explore the `ideas space' --- and to focus on the implementation itself. Furthermore, new members of our research group had to go through different source codes developed by former members.

Extensions of SDM have been used in many applications. For example, \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used a modified SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs. \citet{fan1997genetic} extended SDM with genetic algorithms. \citet{chada2016you} extended SDM creating the Rotational Sparse Distributed Memory (RSDM), which is used to modeling network motifs, dynamic flexibility, and hierarchical organization, all results from neuroscience literature.

@@

An important observation is that around 800,000 neuron deaths (80\% of all neurons) the critical distance becomes small, i.e., SDM recall capacity is very diminished. After 900,000 neuron deaths, the critical distance is zero. In this case, everything has been lost.

@@

We run exactly the same simulation for a 256-bit SDM with 1,000,000 hard locations. The results were even more surprising, as the 256-bit SDM seems to be more robust to loss of neurons than the 1,000-bit SDM (see Figure \ref{fig:sdm-neuron-death-256bits}). Notice that the loss of 50\% of neurons barely affected the 256-bit SDM which remained functional even facing an enormous loss of 90\% of neurons.

@@

The results, however, have not yielded performance improvements. Though for $z \leq 1$ results are comparable to $z=1$, for $z>1$, the system shows a clear deterioration, with a smaller critical distance and faster divergence at large-distance reads. This is shown in Figures \ref{fig:murillo-generalization-experiments} and \ref{fig:murillo-generalization-experiments-6reads}.

@@

An unexplored idea is to use multiple SDMs which communicate. A first SDM would write the whole picture, just like we have done. Another SDM would write specific parts of the image, just like our eye focusing on specific parts. When reading, they may compose the counters and give a more precise classification.

@@

One intriguing question here is:  is there a `better' number of dimensions and of hard locations?  If so, can such numbers be better studied algebraically or numerically?

>>>>

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. This model may be applied in all sort of applications because of its incredible ability to closely reflect the human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where one's fear comes from. One just feels it. We may interpret this situation as clues of the present --- a dark alley; a giant metropolitan area; people going on about their lives mostly indifferent from each other; constrained routes ahead and behind you; a general lack of activity; that very feeling that something isn't right... without knowing precisely what isn't right, or even what `right' means... etc. --- recalling past experiences from memory and thus generating the feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, one's brain involuntarily makes an analogy and recognizes the possibility of danger. This flexibility into mapping one situation in another is an essential human feature which is hard to replicate in computers.

@@

Our motivation was our own effort to run our models. As there is no reference implementation, we had to develop our own and run several simulations to ensure that our implementation was correct and bug-free. Thus, we had to deviate from our primary goal --- which was to test our hypothesis and explore the `ideas space' --- and to focus on the implementation itself. Furthermore, new members of our research group had to go through different source codes developed by former members.

Extensions of SDM have been used in many applications. For example, \citet{Snaider2011} extended SDM to store sequences of vectors and trees efficiently.  \citet{Rajesh1998} used a modified SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs. \citet{fan1997genetic} extended SDM with genetic algorithms. \citet{chada2016you} extended SDM creating the Rotational Sparse Distributed Memory (RSDM), which is used to modeling network motifs, dynamic flexibility, and hierarchical organization, all results from neuroscience literature.

@@

An important observation is that around 800,000 neuron deaths (80\% of all neurons) the critical distance becomes small, i.e., SDM recall capacity is hugely diminished. After 900,000 neuron deaths, the critical distance is zero. In this case, everything has been lost.

@@

We run the same simulation for a 256-bit SDM with 1,000,000 hard locations. The results were even more surprising, as the 256-bit SDM seems to be more robust to loss of neurons than the 1,000-bit SDM (see Figure \ref{fig:sdm-neuron-death-256bits}). Notice that the loss of 50\% of neurons barely affected the 256-bit SDM which remained functional even facing an enormous loss of 90\% of neurons.

@@

The results, however, have not yielded performance improvements. Though for $z \leq 1$ results are comparable to $z=1$, for $z>1$, the system shows an evident deterioration, with a smaller critical distance and faster divergence at large-distance reads. This is shown in Figures \ref{fig:murillo-generalization-experiments} and \ref{fig:murillo-generalization-experiments-6reads}.

@@

An unexplored idea is to use multiple SDMs which communicate. A first SDM would write the whole picture, just like we have done. Another SDM would write specific regions of the image, just like our eye focusing on specific regions. When reading, they may compose the counters and give a more precise classification.

@@

One intriguing question here is:  is there a `better' number of dimensions and hard locations?  If so, can such numbers be better studied algebraically or numerically?

### 2018-03-16 Chapters/sdm-applications.tex
SDM more grammar and typo fixes

Supervised classification problem consists of categorizing data into groups after seeing some samples from each group. First, it is presented pieces of data with their categories. The algorithm learns from these data, which is known as the learning phase. Then, new pieces of data are presented and the algorithm must classify them into the already known groups. It is named supervised because  the algorithm will not create the groups itself. It will learn the groups from during the learning phase, in which the groups have already been defined and the pieces of data have already been classified into them.

@@

The images had 31 pixels of width and 32 pixels of height, totaling 992 pixels per image. Each image was mapped into a 1,000-bit bitstring in which the bits were set according to the color of each pixel of the image. So, white pixels were equal to bit 0, and black pixels to bit 1. The 8 remaining bits were all set to zero. This was a bijective mapping (or one-to-one mapping), i.e., there was only one bitstring for each image, and there was only one image for each bitstring.

@@

Finally, we have assessed the performance of our classifier. We had done it in three different scenarios: high noise (20\%), low noise (5\%) and no noise. See Figures \ref{fig-classification-noise-high} and \ref{fig-classification-no-noise} for images with 20\% noise and no noise. The low noise scenario had the same noise as the training set. For each scenario, we had classified 620 unknown images with 10 images per group.

@@

In the no noise scenario, the classifier has hit all characters, except letter ``l'' which was wrongly associated with the group of ``i''. We believe that it happened because the classifier had never seen an image with no noise and the difference between the images of ``l'' and ``i'' is smaller than the critical distance. So, both groups have been merged and it would converge to only one of them. In our simulation, it happened to be the group of ``i''.

@@

The critical distance plays an important role in the classification error. As we have 62 groups and each has been trained with 100 images, there were 6,200 writes to the memory. When an image is being classified, it will have to converge to a group, and the convergence depends on the distance between this image and the images from the training set, i.e, in the noise level.

@@

These results show that the SDM may be used as a supervised classification algorithm. Although we do not believe that the mapping between images and bitstrings are even close to the way human cognition deals with images, we believe the results are interesting and useful to many possible real-world problems.

@@

Image noise filtering consists in removing the noise from an input, in our case an image. Our images are black \& white images and the noise is generated randomly flipping some of their pixels from black to while and vice versa. In Figure \ref{fig-filter-progressive-noise}, we may see an image with different levels of noise, from 0\% to 45\% in steps of 5\%. It makes no sense to apply 50\% of noise because it would absolutely randomize the image.

@@

The images have 30 x 30 pixels, totaling 900 pixels per image. Each image is mapped into a bitstring of 1,000 bit in which the bits are set according to the color of each pixel of the image. White pixels are mapped to bit 0, and black pixels to bit 1. The 100 remaining bits are all set to zero. This is a bijective mapping (or one-to-one) from images and bitstrings, i.e., there is one, and only one, bitstring for each image, and vice versa.

@@

Then, in order to test the filtering, we just have read from noisy images, and the results were really impressive. We were able to clean images up to 42\% of noise (see Figure \ref{fig:filter-testing}). While SDM has never seen a clean version of the letters, it just learned from the learning phase which pixels have appeared more frequently and choose them.

@@

If the intersection between images becomes too high, the noise filter stops working properly. We have confirmed it writing the letters ``B'', ``C'', ``D'' and the numeral ``8''. They share a great number of hard locations and our noise filter could not filter their noise correctly. The training sets can be seen in Figure \ref{fig:filter-training2} and the results in Figure \ref{fig:filter-results2}.

@@

Our algorithm to decide what should be played is very simple: it reads the current board from SDM and then it chooses the valid action with the highest score. To calculate the scores, the bitstring is split into 9 parts, one per action (Figure \ref{fig:ttt-action-mapping}). The number of bits 1 in each part indicates the score of its respective action.

@@

There are also weights associated with positive, negative, and neutral rewards. They are used to indicate what goal is more important. For instance, if SDM should try to win in the first place, no matter if it may lead to losing, the weight of the positive reward should be higher than the others. But, if it is more important to not lose, then the weight of the negative reward should be higher.

In the very beginning, SDM is empty and its counters are zeroed. So, any reading will result in a random bitstring, because it will flip a coin for all counters. Thus, the chosen action will be random, as their scores will follow a binomial distribution. It is exactly the desired behavior --- we play randomly until we learn. In fact, it will happen every time an unknown board is seen.

@@

When playing against the random player, SDM has already started improving after the first 100 training games. Its winning rate converges quickly to around 80\%, while the drawing rate starts to grow after 2,000 games and keeps growing until the end. The losing rate keeps decreasing until it reaches cycles of 100 games without any loss. See Figure \ref{fig:ttt-results-random}.

When playing against the smart player, SDM has started learning how to not lose during the first training cycles. The drawing rate grows quickly in the first 500 games and slowly since then. The winning rate grew to around 20\% and remained there until 6,000 games, while the drawing rate kept growing. Then, after learning how to not lose, SDM started learning how to win, since, after 7,000 games, the winning rate started to increase. See Figure \ref{fig:ttt-results-smart}.

When playing against another SDM player, both player quickly learns how to not lose. During the first 100 testing games, without any learning game, they have behaved like two random players. After the 100 training games, they have learned how to not lose and the drawing rate grows quickly, reaching 100\%. See Figure \ref{fig:ttt-results-sdm}.

>>>>

Supervised classification problem consists of categorizing data into groups after seeing some samples from each group. First, it is presented pieces of data with their categories. The algorithm learns from these data, which is known as the learning phase. Then, new pieces of data are presented and the algorithm must classify them into the already known groups. It is named ``supervised'' because  the algorithm will not create the groups itself. It will learn the groups during the learning phase, in which the groups have already been defined and the pieces of data have already been classified into them.

@@

The images had 31 pixels of width and 32 pixels of height, totaling 992 pixels per image. Each image was mapped into a 1,000-bit bitstring in which the bits were set according to the color of each pixel of the image. So, white pixels were equal to bit 0, and black pixels to bit 1. The eight remaining bits were all set to zero. This was a bijective mapping (or one-to-one mapping), i.e., there was only one bitstring for each image, and there was only one image for each bitstring.

@@

Finally, we have assessed the performance of our classifier. We had done it in three different scenarios: high noise (20\%), low noise (5\%) and no noise. See Figures \ref{fig-classification-noise-high} and \ref{fig-classification-no-noise} for images with 20\% noise and no noise. The low noise scenario had the same noise as the training set. For each scenario, we had classified 620 unknown images with ten images per group.

@@

In the no noise scenario, the classifier has hit all characters, except letter ``l'' which was wrongly associated with the group of ``i''. We believe that it happened because the classifier had never seen an image with no noise and the difference between the images of ``l'' and ``i'' is smaller than the critical distance. So, both groups have been merged and it converged to only one of them. In our simulation, it happened to be the group of ``i''.

@@

The critical distance plays an important role in the classification error. As we have 62 groups and each has been trained with 100 images, there were 6,200 writes to the memory. When an image is being classified, it will have to converge to a group, and the convergence depends on the distance between this image and the images from the training set, i.e., in the noise level.

@@

These results show that the SDM may be used as a supervised classification algorithm. Although we do not believe that the mapping between images and bitstrings are even close to the way human cognition deals with images, we believe the results are exciting and useful to many possible real-world problems.

@@

Image noise filtering consists in removing the noise from an input, in our case an image. Our images are black \& white images, and the noise is generated randomly flipping some of their pixels from black to while and vice versa. In Figure \ref{fig-filter-progressive-noise}, we may see an image with different levels of noise, from 0\% to 45\% in steps of 5\%. It makes no sense to apply 50\% of noise because it would absolutely randomize the image.

@@

The images have 30 x 30 pixels, totaling 900 pixels per image. Each image is mapped into a bitstring of 1,000 bit in which the bits are set according to the color of each pixel of the image. White pixels are assigned to bit 0, and black pixels to bit 1. The 100 remaining bits are all set to zero. This is a bijective mapping (or one-to-one) from images and bitstrings, i.e., there is one, and only one, bitstring for each image, and vice versa.

@@

Then, in order to test the filtering, we just have read from noisy images, and the results were remarkable. We were able to clean images up to 42\% of noise (see Figure \ref{fig:filter-testing}). While SDM has never seen a clean version of the letters, it just learned from the learning phase which pixels have appeared more frequently and choose them.

@@

If the intersection between images becomes too high, the noise filter stops working properly. We have confirmed it writing the letters ``B'', ``C'', ``D'' and the numeral ``8''. They share a high number of hard locations, and our noise filter could not filter their noise correctly. The training sets can be seen in Figure \ref{fig:filter-training2} and the results in Figure \ref{fig:filter-results2}.

@@

Our algorithm to decide what should be played is simple: it reads the current board from SDM and then it chooses the valid action with the highest score. To calculate the scores, the bitstring is split into 9 parts, one per action (Figure \ref{fig:ttt-action-mapping}). The number of bits 1 in each part indicates the score of its respective action.

@@

There are also weights associated with positive, negative, and neutral rewards. They are used to indicate what goal is more important. For instance, if SDM should try to win in the first place, no matter if it may lead to losing, the weight of the positive reward should be higher than the others. But, if it is more important not to lose, then the weight of the negative reward should be higher.

In the very beginning, SDM is empty, and its counters are zeroed. So, any reading will result in a random bitstring, because it will flip a coin for all counters. Thus, the chosen action will be random, as their scores will follow a binomial distribution. It is precisely the desired behavior --- we play randomly until we learn. In fact, it will happen every time an unknown board is seen.

@@

When playing against the random player, SDM has already started improving after the first 100 training games. Its winning rate converges quickly to around 80\%, while the drawing rate starts to grow after 2,000 games and keeps rising until the end. The losing rate keeps decreasing until it reaches cycles of 100 games without any loss. See Figure \ref{fig:ttt-results-random}.

When playing against the smart player, SDM has started learning how not to lose during the first training cycles. The drawing rate grows quickly in the first 500 games and slowly since then. The winning rate grew to around 20\% and remained there until 6,000 games, while the drawing rate kept increasing. Then, after learning how not to lose, SDM started learning how to win, since, after 7,000 games, the winning rate started to increase. See Figure \ref{fig:ttt-results-smart}.

When playing against another SDM player, both player quickly learns how not to lose. During the first 100 testing games, without any learning game, they have behaved like two random players. After the 100 training games, they have learned how not to lose, and the drawing rate grows quickly, reaching 100\%. See Figure \ref{fig:ttt-results-sdm}.

### 2018-03-16 Chapters/sdm-architecture.tex
SDM more grammar and typo fixes

Scanning for the activated hard locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(H)$. The algorithm which seems to best fit in our problem consumes $O(H)$ space and runs in $O(\log^n(H))$ \citep{chazelle1988functional}, which is really slower than $O(H)$ when, for instance, $H=1,000,000$ and $n=1,000$. For a review of the range reporting algorithms, see \citet{chan2011orthogonal}.

In 2014, there was published a solution to fast search in hamming space which seems applicable to our problem \citet{norouzi2014fast}. It provides a fast search when $r/n < 0.11$ or $r/n < 0.06$, where $r$ is the radius and $n$ is the number of bits. But, in our case, for a 1,000 bits SDM, $r/n = 0.451$, which changes the runtime to $O(H^{0.993})$. This is really close to $O(H)$, but with a larger constant. Unfortunately, $O(H)$ is still faster.

@@

There are 8 OpenCL kernels which explore differently the GPU architecture to improve performance. It is necessary because there are several GPU microarchitectures and a single kernel will never be optimal for all of them. In simplified form, OpenCL splits the tasks into workgroups which, in turn, split their part of the task into workers. The works are like threads in a computer. OpenCL specifies four levels of memory hierarchy for the GPUs: global memory, read-only memory, local memory, and private memory. The global memory and read-only memory are accessible by all workgroups, while each workgroup has its own local memory, accessible by its workers. Finally, each worker has its own private memory. The number of workers per workgroup is defined by user and must be multiple of the number of tasks.

All 8 kernels do the same thing: calculate the exclusive OR (XOR) between two 64-bit integers and count the number of bits one in the result. They just do it with different approaches. For instance, \lstinline{single_scan0} calculates one distance between bitstrings per worker (Listing \ref{lst:single_scan0}); while \lstinline{single_scan2} uses a whole workgroup to calculate each distance, distributing each element of the 64-int integer array per worker (Listing \ref{lst:single_scan2}).

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan0},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan0}}. It calculates one distance per worker and let the GPU decide how to distribute this task between workgroups and workers. It is the most simple kernel and does not explore any details of the GPU architecture.},
]
__kernel
void single_scan0(
		__constant const uchar *bitcount_table,
		__global const ulong *bitstrings,
		const uint bs_len,
		const uint sample,
		__constant const ulong *bs,
		const uint radius,
		__global uint *counter,
		__global uint *selected,
		__local uint *partial_dist)
{
	uint id = get_global_id(0);

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan1},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan1}}. It is just like \lstinline{single_scan0}, but it distribute several distances per workgroup, which, in turn, distribute the distances among their workers.},
]
__kernel
void single_scan1(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint id;
    ulong a;
    uint dist;
    const __global ulong *row;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan3},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan3}}. It calculates one distance per workgroup, distributing each 64-bit integer operation per worker, and then summing the results obtained by the workers. The sum algorithm is a parallel reduction, in which the workers split the array in two parts and sum the second part in the first part every loop. So, the sum is calculated in {$O(\log_2(\text{number of workers per workgroup}))$}. This kernel only works when the number of workers per workgroup is a power-of-2.},
]
__kernel
void single_scan3(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan5},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan5}}. This kernel is just like \lstinline{single_scan3}, but it explores one more detail of many GPU microarchitecture: the size of the warp. As the workers in the same warp are always synchronized, there is no need to synchronize them using a barrier. This specific kernel only works when the number of workers per workgroup is a power-of-2.},
]
__kernel
void single_scan5(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
basicstyle=\scriptsize\ttfamily,
label={lst:single_scan6},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan6}}. This kernel is just like \lstinline{single_scan5}, but it works with any number of workers per work. The tradeoff is an aditional step in the parallel reduction algorithm.},
]
__kernel
void single_scan6(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

The framework has both Kanerva's read and Murilo's generalized read. The generalization brings a parameter $z$, which is the exponent. In this case, the results are floating points instead of integers, which considerably reduces performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is another special read operation: the weighted reading. In the weighted reading, the value of the counters is multiplied by a weight which depends only on the distance between the reading address and the hard location address. The weight is retrieved from a lookup table of integers indexed by the distance. The rest of the read operation is exactly the same.

>>>>

Scanning for the activated hard locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(H)$. The algorithm which seems to best fit in our problem consumes $O(H)$ space and runs in $O(\log^n(H))$ \citep{chazelle1988functional}, which is significantly slower than $O(H)$ when, for instance, $H=1,000,000$ and $n=1,000$. For a review of the range reporting algorithms, see \citet{chan2011orthogonal}.

In 2014, there was published a solution to fast search in hamming space which seems applicable to our problem \citet{norouzi2014fast}. It provides a fast search when $r/n < 0.11$ or $r/n < 0.06$, where $r$ is the radius and $n$ is the number of bits. But, in our case, for a 1,000 bits SDM, $r/n = 0.451$, which changes the runtime to $O(H^{0.993})$. This is very close to $O(H)$, but with a larger constant. Unfortunately, $O(H)$ is still faster.

@@

There are eight OpenCL kernels which explore differently the GPU architecture to improve performance. It is necessary because there are several GPU microarchitectures and a single kernel will never be optimal for all of them. In simplified form, OpenCL splits the tasks into workgroups which, in turn, split their part of the task into workers. The works are like threads in a computer. OpenCL specifies four levels of memory hierarchy for the GPUs: global memory, read-only memory, local memory, and private memory. The global memory and read-only memory are accessible by all workgroups, while each workgroup has its own local memory, accessible by its workers. Finally, each worker has its own private memory. The number of workers per workgroup is defined by user and must be multiple of the number of tasks.

All eight kernels do the same thing: calculate the exclusive OR (XOR) between two 64-bit integers and count the number of bits one in the result. They just do it with different approaches. For instance, \lstinline{single_scan0} calculates one distance between bitstrings per worker (Listing \ref{lst:single_scan0}); while \lstinline{single_scan2} uses a whole workgroup to calculate each distance, distributing each element of the 64-int integer array per worker (Listing \ref{lst:single_scan2}).

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan0},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan0}}. It calculates one distance per worker and let the GPU decide how to distribute this task between workgroups and workers. It is the most straightforward kernel and does not explore any details of the GPU architecture.},
]
__kernel
void single_scan0(
		__constant const uchar *bitcount_table,
		__global const ulong *bitstrings,
		const uint bs_len,
		const uint sample,
		__constant const ulong *bs,
		const uint radius,
		__global uint *counter,
		__global uint *selected,
		__local uint *partial_dist)
{
	uint id = get_global_id(0);

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan1},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan1}}. It is just like \lstinline{single_scan0}, but it distributes several distances per workgroup, which, in turn, distributes the distances among their workers.},
]
__kernel
void single_scan1(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint id;
    ulong a;
    uint dist;
    const __global ulong *row;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan3},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan3}}. It calculates one distance per workgroup, distributing each 64-bit integer operation per worker, and then summing the results obtained by the workers. The sum algorithm is a parallel reduction, in which the workers split the array into two parts and sum the second part in the first part every loop. So, the sum is calculated in {$O(\log_2(\text{number of workers per workgroup}))$}. This kernel only works when the number of workers per workgroup is a power-of-2.},
]
__kernel
void single_scan3(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
label={lst:single_scan5},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan5}}. This kernel is just like \lstinline{single_scan3}, but it explores one more detail of many GPU microarchitecture: the size of the warp. As the workers in the same warp are always synchronized, there is no need to sync them using a barrier. This specific kernel only works when the number of workers per workgroup is a power-of-2.},
]
__kernel
void single_scan5(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

\begin{figure}[!p]
\begin{lstlisting}[
language={[OpenCL]C},
basicstyle=\scriptsize\ttfamily,
label={lst:single_scan6},
caption={OpenCL kernel {\lstinline[columns=fixed]{single_scan6}}. This kernel is just like \lstinline{single_scan5}, but it works with any number of workers per work. The tradeoff is an additional step in the parallel reduction algorithm.},
]
__kernel
void single_scan6(
        __constant const uchar *bitcount_table,
        __global const ulong *bitstrings,
        const uint bs_len,
        const uint sample,
        __constant const ulong *bs,
        const uint radius,
        __global uint *counter,
        __global uint *selected,
        __local uint *partial_dist)
{
    uint dist;
    ulong a;
    uint j;

@@

The framework has both Kanerva's read and Murilo's generalized read. The generalization brings a parameter $z$, which is the exponent. In this case, the results are floating points instead of integers, which considerably reduces performance. When $z=1$, it is precisely as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is another particular read operation: the weighted reading. In the weighted reading, the value of the counters is multiplied by a weight which depends only on the distance between the reading address and the hard location address. The weight is retrieved from a lookup table of integers indexed by the distance. The remaining of the read operation is just the same.

### 2018-03-16 Chapters/sdm-critical-distance.tex
SDM more grammar and typo fixes

\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{fig:sdm-10000w-table-7-2}}
\end{figure}

@@

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Figure \ref{fig:sdm-100w-table-7-2-10samples}). This result strengthens our hypothesis that the disparities in the computational results are due to the interaction effect of high numbers of different attractors. In Figure \ref{fig:sdm-table-7-2-steps} we can notice that the more random bitstrings are written, the stronger the attractors.

@@

\caption{Results generated by the framework similar to Kanerva's original Table 7.2. Here we have a 1,000 bit, 1,000,000 hard location SDM with (a) exactly 100 random bitstrings written into it and (b) steps of 1,000 random bitstrings written into it.
\label{fig:sdm-100w-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=\textwidth]{images02/sdm-10000w-table-7-2-15iter.png}
\caption{This graph shows the interaction effects more clearly.  As we change the single read to a 6-iterative read, the effect has vanished and all bitstrings above $x=500$ have converged to 500-bit distance bitstrings. Here we have the exact same configuration of Figure \ref{fig:sdm-10000w-table-7-2}, except for the iterative read.
\label{fig:sdm-10000w-table-7-2-15iter}}
\end{figure}

Going further in the analysis, we calculated the probability of missing a bit when reading from SDM. After all, that's how Kanerva has originally found the curve. To do this, we used the following equations from our previous work \citep{brogliato2014sparse}. Let $d$ be the distance to the target, $h$ be the number of hard locations activated during read and write operations, $s$ be the number of total stored bitstrings, $H$ be the number of total hard locations, $w$ be the number of times the target was written into SDM, $\theta$ be the total random bitstrings in all $h$ hard locations activated by read operation, and $\phi(d)$ be the average number of shared hard locations activated two bitstrings $d$ bits away.

@@

As we can easily notice, we got exactly the same curve as Kanerva. This question has been intriguing us since then and we are still looking for a more analytic explanation than merely interference from the other written attractors.

>>>>

\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard location SDM with 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{fig:sdm-10000w-table-7-2}}
\end{figure}

@@

However, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Figure \ref{fig:sdm-100w-table-7-2-10samples}). This result strengthens our hypothesis that the disparities in the computational outcomes are due to the interaction effect of high numbers of different attractors. In Figure \ref{fig:sdm-table-7-2-steps} we can notice that the more random bitstrings are written, the stronger the attractors.

@@

\caption{Results generated by the framework similar to Kanerva's original Table 7.2. Here we have a 1,000 bit, 1,000,000 hard location SDM with (a) just 100 random bitstrings written into it and (b) steps of 1,000 random bitstrings written into it.
\label{fig:sdm-100w-table-7-2}}
\end{figure}

@@

\begin{figure}[h]
\centering\includegraphics[width=\textwidth]{images02/sdm-10000w-table-7-2-15iter.png}
\caption{This graph shows the interaction effects more clearly.  As we change the single read to a 6-iterative read, the effect has vanished, and all bitstrings above $x=500$ have converged to 500-bit distance bitstrings. Here we have precisely the same configuration of Figure \ref{fig:sdm-10000w-table-7-2}, except for the iterative read.
\label{fig:sdm-10000w-table-7-2-15iter}}
\end{figure}

Going further in the analysis, we calculated the probability of missing a bit when reading from SDM. After all, that is how Kanerva has originally found the curve. To do this, we used the following equations from our previous work \citep{brogliato2014sparse}. Let $d$ be the distance to the target, $h$ be the number of hard locations activated during reading and write operations, $s$ be the number of total stored bitstrings, $H$ be the number of total hard locations, $w$ be the number of times the target was written into SDM, $\theta$ be the total random bitstrings in all $h$ hard locations activated by read operation, and $\phi(d)$ be the average number of shared hard locations activated two bitstrings $d$ bits away.

@@

As we can easily notice, we got exactly the same curve as Kanerva. This question has been intriguing us since then, and we are still looking for a more analytic explanation than merely interference from the other written attractors.
